An Efficient Minibatch Acceptance Test for MCMC

### 2016-07-06 nips_2016.tex
updates to first section

Assume we begin with the current sample $\theta$  and a
candidate sample $\theta'$. Let $u$ be a uniform random variable $u \sim
\mathcal{U}[0,1]$. We accept $\theta'$ if $g(\Delta(\theta,\theta')) > u$, and reject otherwise.
Since $g(s)$ is monotonically increasing, its inverse $g^{-1}(s)$ is well-defined and
unique. So an equivalent test is to accept $\theta'$ iff
\begin{equation}
  \Delta(\theta,\theta') > v = g^{-1}(u)
\end{equation}
where $v$ is a random variable with the logistic distribution (its CDF is the logistic function). To
see this notice that $\frac{du}{dv} = g'$, that $g'$ is the density corresponding to a logistic
CDF, and finally that $\frac{du}{dv}$ is the density of $v$. The density of $v$ is symmetric,
so we can equivalently test whether
\begin{equation}
  \Delta(\theta,\theta') + v > 0
\end{equation}
for a logistic random $v$.

@@

We now move to acceptance testing using the minibatch estimator
$\Delta^*(x,\theta,\theta')$. From equation (\ref{eq:delta3}),
$\Delta^*(x,\theta,\theta')$ can be represented as a constant term
plus a sum of $b$ iid terms of the form
$\log\frac{p(x_i|\theta')}{p(x_i|\theta)}$. As $N$ increases, $\Delta^*(\theta,\theta')$
therefore has a distribution which approaches a normal distribution by the central limit
theorem. Later we will provide specific bounds between the CDF of $\Delta^*(\theta,\theta')$
and a Normal CDF, but for now we use an aymptotic argument.

\begin{lemma}\label{lem:gaussian}
If the minibatch data is chosen randomly without replacement and has sufficiently many elements,
then the distribution of $\Delta'$ is approximately Gaussian.
\end{lemma}

\begin{proof}
By expanding the definition of $\Delta'$ from Equation~\ref{eq:deltas}, we get
\[
\Delta' = \log p(\theta') - \log p(\theta) + \log q(\theta \mid \theta') - \log q(\theta' \mid \theta) +
\frac{N}{n}\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta)).
\]
During a given iteration, $\theta$ and $\theta'$ are fixed. Therefore, the randomness in $\Delta'$
comes only from the minibatch. Since the minibatch is chosen randomly, the distribution of
$\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta))$ converges to a Gaussian for
sufficiently large $N$ by the Central Limit Theorem.
\end{proof}

% Daniel: I decided to put the figure here, so it's closer to the text that references it.
\begin{figure}[t]
    \centering
    \includegraphics[width=1\textwidth]{mh_convolution_diagram_v2}
    \caption{
    Three examples of $X_{\rm norm}$ and $X_{\rm corr}$ distributions that convolve to form the
    standard logistic distribution. We use three standard deviation values of $X_{\rm norm}$. The
    two red (respectively, blue, green) curves form the logistic. The $y$-axis is capped at 0.5.
    This figure must be viewed in color.
    }
    \label{fig:deconvolution}
    %\vspace{-10pt}
\end{figure}

From Lemma~\ref{lem:gaussian}, since $\Delta'$ is a noisy approximation of $\Delta$, the
relationship is expressed as
\begin{equation}\label{eq:relationship}
\Delta' = \Delta + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2),
\end{equation}
for a Gaussian noise term $\varepsilon$.  The noise means we cannot just replace $\Delta+X$ with
$\Delta'+X$ for our test, so we need to slightly change our acceptance criteria. Our insight is to
decompose $X$ as
\begin{equation}\label{eq:deconvolution}
X = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ is a zero-mean Gaussian and $X_{\rm corr}$ is a zero-mean
``correction'' term.  These two add (i.e., their distributions convolve) to form $X$. The criteria
to accept is now
\begin{equation}\label{eq:criteria}
\Delta + X = \Delta + X_{\rm norm} + X_{\rm corr} \approx \Delta' + X_{\rm corr} >0.
\end{equation}
The approximation exists because $X_{\rm norm}$ is an estimate of $\varepsilon$. If $X_{\rm norm}$
is exact, then we get the same MH test with $\Delta'$ as we do with the full data version $\Delta$.

>>>>

Assume we begin with the current sample $\theta$  and a
candidate sample $\theta'$. Let $V$ be a uniform random variable $V \sim
\mathcal{U}(0,1)$. We accept $\theta'$ if $g(\Delta(\theta,\theta')) > V$, and reject otherwise.
Since $g(s)$ is monotonically increasing, its inverse $g^{-1}(s)$ is well-defined and
unique. So an equivalent test is to accept $\theta'$ iff
\begin{equation}
  \Delta(\theta,\theta') > X = g^{-1}(V)
\end{equation}
where $X$ is a random variable with the logistic distribution (its CDF is the logistic function). To
see this notice that $\frac{dV}{dX} = g'$, that $g'$ is the density corresponding to a logistic
CDF, and finally that $\frac{dV}{dX}$ is the density of $X$. The density of $X$ is symmetric,
so we can equivalently test whether
\begin{equation}
  \label{theexacttest}
  \Delta(\theta,\theta') + X > 0
\end{equation}
for a logistic random $X$.

@@

We now move to acceptance testing using the minibatch estimator
$\Delta^*(x,\theta,\theta')$. From equation (\ref{eq:delta3}),
$\Delta^*(x,\theta,\theta')$ can be represented as a constant term
plus a sum of $b$ iid terms of the form
$\log\frac{p(x_i|\theta')}{p(x_i|\theta)}$. As $N$ increases, $\Delta^*(\theta,\theta')$
therefore has a distribution which approaches a normal distribution by the central limit
theorem. Later we will provide specific bounds between the CDF of $\Delta^*(\theta,\theta')$
and a Normal CDF, but for now we use an asymptotic argument.

In the limit, since $\Delta^*$ is normally distributed about its mean $\Delta$, we
can write
\begin{equation}\label{eq:relationship}
\Delta^* = \Delta + X_{\rm norm}, \quad X_{\rm norm} \sim \mathcal{\bar{N}}(0, \sigma^2(\Delta^*)),
\end{equation}
where $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ denotes a distribution which is approximately
normal with variance $\sigma^2(\Delta^*)$. 
But to perform the test in equation (\ref{theexacttest}) we want $\Delta + X$ for
a logistic random $X$ (call it $X_{\rm logistic}$ from now on). 
Our approach is to decompose $X_{\rm logistic}$ as
\begin{equation}\label{eq:deconvolution}
X_{\rm logistic} = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ is a zero-mean Gaussian and $X_{\rm corr}$ is a zero-mean
``correction'' term.  These two add (i.e., their distributions convolve) to form $X_{\rm logistic}$. The acceptance
test is now:
\begin{equation}\label{eq:criteria}
\Delta + X_{\rm logistic} = (\Delta + X_{\rm norm}) + X_{\rm corr} = \Delta^* + X_{\rm corr} >0.
\end{equation}
If $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ is exactly $\mathcal{N}(0, \sigma^2(\Delta^*))$,
the above test is exact as well. And as we will see later,  if there is a maximum error $\epsilon$ between
the CDF of $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ and the CDF of $\mathcal{N}(0, \sigma^2(\Delta^*))$,
then the acceptance test has an error of at most $\epsilon$.
